Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers

### 2018-03-08 paper/Final/sparse_poly_icml.tex
fixed a lot of capitalization of K issues, removed some sentences, clarified some things.

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Speeding Up Polynomial Feature Expansions of Compressed Sparse Row Matrices}

@@

\twocolumn[
\icmltitle{Speeding Up Polynomial Feature Expansions by Leveraging Density}

@@

\begin{icmlauthorlist}
\icmlauthor{Andrew Nystrom}{google}
\icmlauthor{John Hughes}{brown}
\end{icmlauthorlist}

\icmlaffiliation{google}{Google Inc.}
\icmlaffiliation{brown}{Brown University}

\icmlcorrespondingauthor{Andrew Nystrom}{nystrom@google.com}
\icmlcorrespondingauthor{John Hughes}{jfh@cs.brown.edu}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boring formatting information, machine learning, ICML}

@@

@@

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
%\footnotetext{hi}

\begin{abstract}
We provide an algorithm that speeds up polynomial and interaction feature expansion on sparse matrices.
The algorithm operates on and produces a compressed sparse row matrix with no densification.
It works by mapping (via simplex numbers) column indices with nonzero elements to a column index in the polynomial expansion vector, allowing for only the nonzero elements in a CSR matrix to be iterated over when computing products.
For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
\end{abstract}

@@

A well known and widely used feature mapping is the \emph{polynomial expansion}, which produces a new feature for each degree-$k$ monomial in the original features. (If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy$, and the order-3 ones are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.)
A $k$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
This mapping was first utilized in a published experiment by Joseph Diaz Gergonne in 1815 \cite{gergonne1974application, smith1918standard}.

@@

Uppercase $K$ refers to the degree of a polynomial or interaction expansion.
When used as a superscript, $K$ indicates that element below it is in a polynomially expanded context.
For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^K$ is the number of nonzero elements in the polynomial expansion of that row vector.
If lowercase $k$ refers to a column index.

@@

@@

It should be noted that in practice, as well as in our code and experiments, expansions for degrees $1, 2, \dots, k-1$ are also generated.
The final design matrix is the augmentation of all such expansions.
However, the algorithm descriptions in this paper omit these steps as they would become unnecessarily visually and notationally cumbersome.
Extending them to include all degrees less than $K$ is trivial and does not affect the complexity of the algorithm as the terms that involve $k$ dominate.

@@

To develop a general algorithm, we require a mapping from a list of $k$ columns of $\bm{A}$ to a column of $\bm{A}^k$.
If there are $D$ columns of $\bm{A}$ and $\binom{D}{K}$ columns of $\bm{A}^K$, this can be accomplished by a bijective mapping of the following form:

@@

Stated more verbosely, we require a bijective mapping from tuples consisting of column indicies of the original input to where the column index of the corresponding product of features in the polynomial expansion.
While any bijective mapping would suffice, a common order in which to produce polynomial features is $(0, 1), (0, 2), \dots, (0, D-1), (1, 1), (1, 2), \dots (1, D-1), \dots (D-1, D-1)$ for $K=2$ where the elements of the tuples are column indices.
That is, the further to the right an index is, the sooner it is incremented.
We refer to this order as \emph{big endian order} as it can be thought of as the way digits in a big endian number system are incremented when counting.
If we want for our algorithm to be backwards compatible with existing models, the mapping must use the big endian ordering.

@@

For reference, we list the first four triangle and tetrahedral numbers in the following table:

@@

Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
The complexity of the algorithm, for fixed $k \ll dD$, is therefore
\begin{align}
& \Theta\left(\binom{dD+K-1}{k}\right) = \Theta\left(\frac{(dD+K-1)!}{K!(dD-1)!}\right)\\
& = \Theta\left(\frac{(dD+K-1)(dD+K-2) \dots (dD)}{K!}\right)\\
& = \Theta\left((dD+K-1)(dD+K-2) \dots (dD)\right)\\
& = \Theta\left(d^KD^K\right)
\end{align}

For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^KD^K\right)$.
The dense algorithm does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.

\subsection{Empirical Results}
To verify the average time complexity of our algorithm, we implemented both the sparse version and the baseline in the Cython programming language so that results would be directly comparable.
We sought the relationships between runtime and the instance count ($N$), the instance dimensionality ($D$), and the instance density ($d$).

>>>>

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers}

@@

\twocolumn[

@@

\author{
  Nystrom, Andrew\\
  \texttt{nystrom@google.com}
  \and
  Hughes, John\\
  \texttt{jfh@cs.brown.edu}
}
\title{Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers}
\maketitle

@@

\date{}

@@

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
%\footnotetext{hi}

\begin{abstract}
We provide an algorithm that speeds up polynomial and interaction feature expansion on sparse matrices.
The algorithm operates on and produces a compressed sparse row matrix with no densification.
It works by defining a bijective function involving simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
This function allows for only the nonzero elements to be iterated over and multiplied together, greatly improving the expansion of sparse matrices in compressed sparse row format.
For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
\end{abstract}

@@

A well known and widely used feature mapping is the \emph{polynomial expansion}, which produces a new feature for each degree-$k$ monomial in the original features. (If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy$, and the order-3 ones are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.
A $K$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
This mapping was first utilized in a published experiment by Joseph Diaz Gergonne in 1815 \cite{gergonne1974application, smith1918standard}.

@@

@@

Uppercase $K$ refers to the degree of a polynomial or interaction expansion.
When used as a superscript of a non-scalar, $K$ indicates that element below it is in a polynomially expanded context.
For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^K$ is the number of nonzero elements in the polynomial expansion of that row vector.
Lowercase $k$ refers to a column index.

@@

It should be noted that in practice, as well as in our code and experiments, expansions for degrees $1, 2, \dots, k-1$ are also generated.
The final design matrix is the augmentation of all such expansions.
However, the algorithm descriptions in this paper omit these steps as they would become unnecessarily visually and notationally cumbersome.
Extending them to include all degrees less than $K$ is trivial and does not affect the complexity of the algorithm as the terms that involve $K$ dominate.

@@

To develop a general algorithm, we require a mapping from a list of $K$ columns of $\bm{A}$ to a single column of $\bm{A}^K$.
If there are $D$ columns of $\bm{A}$ and $\binom{D}{K}$ columns of $\bm{A}^K$, this can be accomplished by a bijective mapping of the following form:

@@

Stated more verbosely, we require a bijective mapping from tuples consisting of column indicies of the original input to where the column index of the corresponding product of features in the polynomial expansion.
While any bijective mapping would suffice, a common order in which to produce polynomial features is $(0, 1)$, $(0, 2)$, $\dots$, $(0, D-1)$, $(1, 1)$, $(1, 2)$, $\dots$, $(1, D-1)$, $\dots$, $(D-1, D-1)$ for $K=2$ where the elements of the tuples are column indices.
That is, the further to the right an index is, the sooner it is incremented.
If we want for our algorithm to be backwards compatible with existing models, the mapping must use the this same ordering.

@@

For reference, we list the first five triangle and tetrahedral numbers in the following table:

@@

Calculating $K$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{K}$ products.
The complexity of the algorithm, for fixed $K \ll dD$, is therefore
\begin{align}
& \Theta\left(\binom{dD+K-1}{K}\right) = \Theta\left(\frac{(dD+K-1)!}{K!(dD-1)!}\right)\\
& = \Theta\left(\frac{(dD+K-1)(dD+K-2) \dots (dD)}{K!}\right)\\
& = \Theta\left((dD+K-1)(dD+K-2) \dots (dD)\right)\\
& = \Theta\left(d^KD^K\right)
\end{align}

For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^KD^K\right)$.
The dense algorithm (algorithm \ref{alg:Dense-Second-Order-Polynomial-Expansion}) does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.

\subsection{Empirical Results}
To empirically verify the average time complexity of our algorithm, we implemented both the sparse version and the baseline in the Cython programming language so that results would be directly comparable.
We sought the relationships between runtime and the instance count ($N$), the instance dimensionality ($D$), and the instance density ($d$).
